A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-28 sections/algorithm.tex
S3 algorithm. 4 Checking judgments: discussion on Rule 21

\paragraph{\bf 4. Checking judgements.}
Rules 18-21 deal with checking judgments.
Rule 18 is similar to $\mathtt{DeclSub}$, but rewritten in the
continuation-passing-style.
\bruno{The side condition needs to be explained, and some mention of
  the overlapping is probably needed here.}
Rules 19 and 20 adapt their declarative counterparts to the worklist style.
Rule 21 is a special case of $\mathtt{Decl\to I}$.\bruno{Rule 21
  deserves some more explanation. What is the motivation for this
  rule? You may want to give some concrete example to motivate it.}
The same instantiation technique as in Rules 10 and 11 applies.
The declarative checking rule $\mathtt{Decl1I}$ does not have a direct counterpart in the algorithm, 
because Rules 18 and 24 can be combined to give the same result.

@@

Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$
through propagation when existential varaibles are solved or partially solved.
The variable scopes are maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the declaration of the function type ($\al,\bt$) is only visible to
the checking and the remaining chain $[\al\to\bt/a]\jg$.
The recycling process of Rule 25 differs from DK's corresponding rule significantly,
and we further discuss the differences in Section~\ref{sec:discussion:scoping}.

>>>>

\paragraph{\bf 4. Checking judgements.}
Rules 18-21 deal with checking judgments.
Rule 18 is similar to $\mathtt{DeclSub}$, but rewritten in the
continuation-passing-style.
The side conditions $e \neq \lam e'$ and $B \neq \all B'$ is to
prevent overlapping with Rules 19, 20 and 21,
which are further discussed in the last part of this section.
\bruno{The side condition needs to be explained, and some mention of
  the overlapping is probably needed here.}
Rules 19 and 20 adapt their declarative counterparts to the worklist style.
Rule 21 is a special case of $\mathtt{Decl\to I}$,
dealing with the case when the input type is an existential variable,
representing a monotype \emph{function} as in the declarative system
(it must be a function type, since the expression is a function).
\bruno{Rule 21
  deserves some more explanation. What is the motivation for this
  rule? You may want to give some concrete example to motivate it.}
The same instantiation technique as in Rules 10 and 11 applies.
The declarative checking rule $\mathtt{Decl1I}$ does not have a direct counterpart in the algorithm, 
because Rules 18 and 24 can be combined to give the same result.

\paragraph{Design Variant for Rule 21}
A combination of Rules $18'$ (Rule 18 without the restriction on $e$) and 25 and further derivations
may result in a similar effect to Rule 21.
Rule 21 guesses a monotype for the lambda function $\lam e$, so as Rule 25.
The only difference is that a subtyping relation is introduced by Rule $18'$,
and is to be processed after the type inference:
$$\Gm \Vdash \lam e \Lto \al \rrule{18'} \Gm \Vdash \lam e \To_a a \le \al
\rrule{25} \Gm, \bt[1], \bt[2] \Vdash \bt[1] \to \bt[2] \le \al, x:\bt[1] \Vdash e \Lto \bt[2]$$
Since two monotypes with a valid subtyping relation indicates equality,
eventually $\bt[1] \to \bt[2]$ and $\al$ should be unified with each other.
DK's algorithm takes a similar approach to such variant,
and we believe that they are equivalent.

Choosing Rule 21 simplifies the side condition on Rule 18,
the metatheory and the design: whenever an input type could be a function type,
the algorithmic rule should always take care of
the possibility when the type being a single existential variable.
This pattern applies to all the possible cases, including Rules 10, 11, 21 and 29.

@@

Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$
through propagation when existential variables are solved or partially solved.
The variable scopes are maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the declaration of the function type ($\al,\bt$) is only visible to
the checking and the remaining chain $[\al\to\bt/a]\jg$.
The recycling process of Rule 25 differs from DK's corresponding rule significantly,
and we further discuss the differences in Section~\ref{sec:discussion:scoping}.
